Designing an effective intervention of motivational interview using sequence classification methods

### 2017-02-28 paper/AMIA/main.tex
completed abstract writing

\noindent{\bf Abstract}
\textit{Childhood obesity is a serious public health concern in the United States and worldwide. Therefore, there is a need for identification of successful and unsuccessful motivational interviews to facilitate the development of effective interventions for childhood obesity. This paper describes our work on the classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment versus negative change talk or commitment. Markov chain and hidden Markov models were employed and compared with a different order for the classification of utterance sequences. The classification performances achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using 1st order Markov chain and second order hidden Markov model, respectively. }

@@

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which include a total of 11,353 segmented and annotated utterances. A fragment of an adolescent session transcript is presented in Table~\ref{table:anno_examp}. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider} with 115 different behavior codes, which are grouped into the youth, caregiver, and counselor code groups. These utterances are then scanned from first to last and divided into successful and unsuccessful sequences when a commitment or change talk is found. A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one dataset that includes all sequences of positive change talk or commitment language (882 sequences) and the other dataset which includes utterances ended with negative change talk or commitment language (192 sequences). In our experiment, these two subsets are utilized to train two Markov chain or HMM.\\

@@

\subsection*{\textit{Performance of Markov model in conjunction with determining PPC code sequences that are most likely translate into CT}}
Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis. Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}. From Table~\ref{table:sequence}, it follows that our designed markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092. The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}. It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.\\

>>>>

\noindent{\bf Abstract}
\textit{Childhood obesity is a serious public health concern in the United States and worldwide. Therefore, there is a need for identification of successful motivational interviews to facilitate the development of effective interventions for childhood obesity. This paper presents our work on the classification of utterance sequences into successful and unsuccessful behavior outcomes. Markov chain and hidden Markov models were employed and compared with a different order for the classification of utterance sequences. The classification performances achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using 1st order Markov chain and second order hidden Markov model, respectively. This shows that the behavior codes carry some inherited information to distinguish motivational interviews. We expect that the result will improve if we use a more discriminable data set. Furthermore, the results suggest us a new way to think about designing the motivational interview.}

@@

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which include a total of 11,353 segmented and annotated utterances. A fragment of an adolescent session transcript is presented in Table~\ref{table:anno_examp}. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider} with 115 different behavior codes, which are grouped into the youth, caregiver, and counselor code groups. These utterances are then scanned from first to last and divided into successful and unsuccessful sequences when a commitment or change talk is found. A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one dataset that includes all sequences of positive change talk or commitment language (882 sequences) and the other dataset which includes utterances ended with negative change talk or commitment language (192 sequences). In our experiment, these two subsets are utilized to train two Markov chain or HMM.

@@

\subsection*{\textit{Performance of Markov model in conjunction with determining PPC code sequences that are most likely translate into CT}}
Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis. Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}. From Table~\ref{table:sequence}, it follows that our designed markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092. The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}. It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.

Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses. Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences. \\
